Statistical analysis of periodic data in neuroscience

### 2021-07-06 manuscript/manuscript.Rmd
Minor edits following additional reviewer comments (Appendix removed)

where $N_k$ is the number of observations in group _k_, $\bar{x}_k$ is the bivariate sample mean of group _k_, $\bar{x}_{grand}$ is the bivariate mean of the entire sample (including all groups), $\Sigma$ denotes summation across groups, and $df_M$ is the degrees of freedom for the model. The model degrees of freedom for _k_ groups is $df_M = 2(k-1)$, with the factor of two scaling relative to standard univariate ANOVA reflecting the additional degrees of freedom afforded by having two dependent variables.

@@

where $x_{i,k}$ is the $i$th data point from group _k_, and other terms are as described previously. The degrees of freedom for a balanced design are calculated as $df_R = 2((Nk)-k)$, where _N_ is the number of observations per group, and _k_ is the number of groups. A _p_-value can then be determined by comparing the F-ratio calculated from equation \@ref(eq:Fratio) to an F-distribution with $df_M$ and $df_R$ degrees of freedom. Appendix 1 summarises the results of stochastic simulations that confirm the F-distribution in the absence of an effect.

@@

# Appendix 1 - proof by stochastic simulation of asymptotic F-distributions

To confirm that distributions of F-statistics generated by the $ANOVA^2_{circ}$ test have the expected form, simulations were conducted for a range of study designs. A null bivariate data set was generated for _N_ participants and _k_ groups, using normally distributed random numbers with underlying condition means of 0 for both the real and imaginary components. The F-statistic was then calculated using equations \@ref(eq:Fratio), \@ref(eq:MSM) and \@ref(eq:MSR). This procedure was repeated one million times for twenty combinations of _N_ and _k_, and the distributions summarised using histograms. Figure \@ref(fig:fdistributions) shows the results of these simulations (blue shading), for comparison with analytically derived F-distributions of $df_M = 2(k-1)$ and $df_R = 2((Nk)-k)$ degrees of freedom (black lines, generated using the _df_ function in _R_). The correspondence between the stochastic and analytic distributions is excellent, confirming that the test produces the expected null distribution of test statistics.

  pdf('figures/Fdistributions.pdf', bg="transparent", height = 5, width = 9)

  nsims <- nsims*10  # increase to 1 million simulations for these

  samplesizes <- 2^(2:6)
  nconds <- c(3,5,7,9)
  finex <- seq(0,4,length=1000)

  plotlims <- c(0,30,0,4)
  ticklocsx <- seq(0,4,1)    # locations of tick marks on x axis
  ticklocsy <- 0:4    # locations of tick marks on y axis
  plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4])
  axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2)
  title(xlab="F ratio       ", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5)
  title(ylab="Probability density", line=1, cex.lab=1.5)

  Fmatrix <- array(0,dim=c(length(samplesizes),length(nconds),nsims))

  for (s in 1:length(samplesizes)){

    axis(1, at=ticklocsx+(6*(s-1)), tck=0.01, lab=F, lwd=2)
    mtext(text = ticklocsx, side = 1, line=0.2, at=ticklocsx+(6*(s-1)))
    mtext(text = paste('N =',samplesizes[s]), side=3, at=2+(6*(s-1)))

    for (s2 in 1:length(nconds)){

      grouplabels <- as.factor(rep(1:nconds[s2],each=samplesizes[s]))

      for (n in 1:nsims){

        data <- matrix(rnorm(nconds[s2]*2*samplesizes[s]),nrow=nconds[s2]*samplesizes[s],ncol=2)

        simdata <- complex(real=data[,1],imaginary=data[,2])
        output <- anovacirc.test(data.frame(simdata,grouplabels))
        Fmatrix[s,s2,n] <- output$Fratio
      }

      a <- hist(Fmatrix[s,s2,],breaks = 200, plot=FALSE)
      axvals <- a$mids
      ayvals <- 0.95*a$counts/max(a$counts)
      polygon(c(0,axvals,max(axvals))+(6*(s-1)), c(0,ayvals,0)+(4-s2), col=addalpha(nbdtpal[3],alpha=0.5),border=NA)
      dfM <- 2*(nconds[s2]-1)
      dfR <- 2*((samplesizes[s]*nconds[s2]) - nconds[s2])
      fdist <- df(finex,dfM,dfR)
      fdist <- 0.95*fdist/max(fdist)
      lines(finex+(6*(s-1)),(4-s2)+fdist,lwd=3)

    }}

  text(29,3.5,'k = 3')
  text(29,2.5,'k = 5')
  text(29,1.5,'k = 7')
  text(29,0.5,'k = 9')

  dev.off()

```{r fdistributions, fig.cap = 'Null distributions of F-statistics generated by simulation (blue shading) and analytically (black lines) for different numbers of groups (k; rows) and sample sizes (N; columns).', fig.width=9, fig.height=5, fig.align="center", echo=FALSE}

knitr::include_graphics('figures/Fdistributions.pdf')

>>>>

where $N_k$ is the number of observations in group _k_, $\bar{x}_k$ is the bivariate sample mean of group _k_, $\bar{x}_{grand}$ is the bivariate mean of the entire sample (including all groups), $\Sigma$ denotes summation across groups, and $df_M$ is the degrees of freedom for the model. Note that by using the vector distances between points, this equation reduces multivariate data to a single scalar value, in a similar way to equations \@ref(eq:t2eq) and \@ref(eq:t2c). The model degrees of freedom for _k_ groups is $df_M = 2(k-1)$, with the factor of two scaling relative to standard univariate ANOVA reflecting the additional degrees of freedom afforded by having two dependent variables.

@@

where $x_{i,k}$ is the $i$th data point from group _k_, and other terms are as described previously. The degrees of freedom for a balanced design are calculated as $df_R = 2((Nk)-k)$, where _N_ is the number of observations per group, and _k_ is the number of groups. A _p_-value can then be determined by comparing the F-ratio calculated from equation \@ref(eq:Fratio) to an F-distribution with $df_M$ and $df_R$ degrees of freedom.

@@

<!-- Note that the appendix simulations below were added at the request of a reviewer, and then subsequently removed following the request of the same reviewer (and the editor). -->
<!-- # Appendix 1 - proof by stochastic simulation of asymptotic F-distributions -->

<!-- To confirm that distributions of F-statistics generated by the $ANOVA^2_{circ}$ test have the expected form, simulations were conducted for a range of study designs. A null bivariate data set was generated for _N_ participants and _k_ groups, using normally distributed random numbers with underlying condition means of 0 for both the real and imaginary components. The F-statistic was then calculated using equations \@ref(eq:Fratio), \@ref(eq:MSM) and \@ref(eq:MSR). This procedure was repeated one million times for twenty combinations of _N_ and _k_, and the distributions summarised using histograms. Figure \@ref(fig:fdistributions) shows the results of these simulations (blue shading), for comparison with analytically derived F-distributions of $df_M = 2(k-1)$ and $df_R = 2((Nk)-k)$ degrees of freedom (black lines, generated using the _df_ function in _R_). The correspondence between the stochastic and analytic distributions is excellent, confirming that the test produces the expected null distribution of test statistics. -->

<!-- ```{r include=FALSE} -->

<!-- if (buildfigs==1){ -->

<!--   pdf('figures/Fdistributions.pdf', bg="transparent", height = 5, width = 9) -->

<!--   nsims <- nsims*10  # increase to 1 million simulations for these -->

<!--   samplesizes <- 2^(2:6) -->
<!--   nconds <- c(3,5,7,9) -->
<!--   finex <- seq(0,4,length=1000) -->

<!--   plotlims <- c(0,30,0,4) -->
<!--   ticklocsx <- seq(0,4,1)    # locations of tick marks on x axis -->
<!--   ticklocsy <- 0:4    # locations of tick marks on y axis -->
<!--   plot(x=NULL,y=NULL,axes=FALSE, ann=FALSE, xlim=plotlims[1:2], ylim=plotlims[3:4]) -->
<!--   axis(2, at=ticklocsy, tck=0.01, lab=F, lwd=2) -->
<!--   title(xlab="F ratio       ", col.lab=rgb(0,0,0), line=1.5, cex.lab=1.5) -->
<!--   title(ylab="Probability density", line=1, cex.lab=1.5) -->

<!--   Fmatrix <- array(0,dim=c(length(samplesizes),length(nconds),nsims)) -->

<!--   for (s in 1:length(samplesizes)){ -->

<!--     axis(1, at=ticklocsx+(6*(s-1)), tck=0.01, lab=F, lwd=2) -->
<!--     mtext(text = ticklocsx, side = 1, line=0.2, at=ticklocsx+(6*(s-1))) -->
<!--     mtext(text = paste('N =',samplesizes[s]), side=3, at=2+(6*(s-1))) -->

<!--     for (s2 in 1:length(nconds)){ -->

<!--       grouplabels <- as.factor(rep(1:nconds[s2],each=samplesizes[s])) -->

<!--       for (n in 1:nsims){ -->

<!--         data <- matrix(rnorm(nconds[s2]*2*samplesizes[s]),nrow=nconds[s2]*samplesizes[s],ncol=2) -->

<!--         simdata <- complex(real=data[,1],imaginary=data[,2]) -->
<!--         output <- anovacirc.test(data.frame(simdata,grouplabels)) -->
<!--         Fmatrix[s,s2,n] <- output$Fratio -->
<!--       } -->

<!--       a <- hist(Fmatrix[s,s2,],breaks = 200, plot=FALSE) -->
<!--       axvals <- a$mids -->
<!--       ayvals <- 0.95*a$counts/max(a$counts) -->
<!--       polygon(c(0,axvals,max(axvals))+(6*(s-1)), c(0,ayvals,0)+(4-s2), col=addalpha(nbdtpal[3],alpha=0.5),border=NA) -->
<!--       dfM <- 2*(nconds[s2]-1) -->
<!--       dfR <- 2*((samplesizes[s]*nconds[s2]) - nconds[s2]) -->
<!--       fdist <- df(finex,dfM,dfR) -->
<!--       fdist <- 0.95*fdist/max(fdist) -->
<!--       lines(finex+(6*(s-1)),(4-s2)+fdist,lwd=3) -->

<!--     }} -->

<!--   text(29,3.5,'k = 3') -->
<!--   text(29,2.5,'k = 5') -->
<!--   text(29,1.5,'k = 7') -->
<!--   text(29,0.5,'k = 9') -->

<!--   dev.off() -->

<!-- } -->

<!-- ``` -->

<!-- ```{r fdistributions, fig.cap = 'Null distributions of F-statistics generated by simulation (blue shading) and analytically (black lines) for different numbers of groups (k; rows) and sample sizes (N; columns).', fig.width=9, fig.height=5, fig.align="center", echo=FALSE} -->

<!-- knitr::include_graphics('figures/Fdistributions.pdf') -->

<!-- ``` -->

### 2021-07-06 manuscript/manuscript.tex
Minor edits following additional reviewer comments (Appendix removed)

\title{Statistical analysis of periodic data in neuroscience}
\author{Daniel H. Baker}
\date{2021-06-26}

@@

where \(N_k\) is the number of observations in group \emph{k}, \(\bar{x}_k\) is the bivariate sample mean of group \emph{k}, \(\bar{x}_{grand}\) is the bivariate mean of the entire sample (including all groups), \(\Sigma\) denotes summation across groups, and \(df_M\) is the degrees of freedom for the model. The model degrees of freedom for \emph{k} groups is \(df_M = 2(k-1)\), with the factor of two scaling relative to standard univariate ANOVA reflecting the additional degrees of freedom afforded by having two dependent variables.

@@

where \(x_{i,k}\) is the \(i\)th data point from group \emph{k}, and other terms are as described previously. The degrees of freedom for a balanced design are calculated as \(df_R = 2((Nk)-k)\), where \emph{N} is the number of observations per group, and \emph{k} is the number of groups. A \emph{p}-value can then be determined by comparing the F-ratio calculated from equation \eqref{eq:Fratio} to an F-distribution with \(df_M\) and \(df_R\) degrees of freedom. Appendix 1 summarises the results of stochastic simulations that confirm the F-distribution in the absence of an effect.

@@

\hypertarget{appendix-1---proof-by-stochastic-simulation-of-asymptotic-f-distributions}{%
\section{Appendix 1 - proof by stochastic simulation of asymptotic F-distributions}\label{appendix-1---proof-by-stochastic-simulation-of-asymptotic-f-distributions}}

To confirm that distributions of F-statistics generated by the \(ANOVA^2_{circ}\) test have the expected form, simulations were conducted for a range of study designs. A null bivariate data set was generated for \emph{N} participants and \emph{k} groups, using normally distributed random numbers with underlying condition means of 0 for both the real and imaginary components. The F-statistic was then calculated using equations \eqref{eq:Fratio}, \eqref{eq:MSM} and \eqref{eq:MSR}. This procedure was repeated one million times for twenty combinations of \emph{N} and \emph{k}, and the distributions summarised using histograms. Figure \ref{fig:fdistributions} shows the results of these simulations (blue shading), for comparison with analytically derived F-distributions of \(df_M = 2(k-1)\) and \(df_R = 2((Nk)-k)\) degrees of freedom (black lines, generated using the \emph{df} function in \emph{R}). The correspondence between the stochastic and analytic distributions is excellent, confirming that the test produces the expected null distribution of test statistics.

{\centering \includegraphics{figures/Fdistributions}

\caption{Null distributions of F-statistics generated by simulation (blue shading) and analytically (black lines) for different numbers of groups (k; rows) and sample sizes (N; columns).}\label{fig:fdistributions}
\end{figure}

>>>>

\title{Statistical analysis of periodic data in neuroscience}
\author{Daniel H. Baker}
\date{2021-07-06}

@@

where \(N_k\) is the number of observations in group \emph{k}, \(\bar{x}_k\) is the bivariate sample mean of group \emph{k}, \(\bar{x}_{grand}\) is the bivariate mean of the entire sample (including all groups), \(\Sigma\) denotes summation across groups, and \(df_M\) is the degrees of freedom for the model. Note that by using the vector distances between points, this equation reduces multivariate data to a single scalar value, in a similar way to equations \eqref{eq:t2eq} and \eqref{eq:t2c}. The model degrees of freedom for \emph{k} groups is \(df_M = 2(k-1)\), with the factor of two scaling relative to standard univariate ANOVA reflecting the additional degrees of freedom afforded by having two dependent variables.

@@

where \(x_{i,k}\) is the \(i\)th data point from group \emph{k}, and other terms are as described previously. The degrees of freedom for a balanced design are calculated as \(df_R = 2((Nk)-k)\), where \emph{N} is the number of observations per group, and \emph{k} is the number of groups. A \emph{p}-value can then be determined by comparing the F-ratio calculated from equation \eqref{eq:Fratio} to an F-distribution with \(df_M\) and \(df_R\) degrees of freedom.

@@
